This section presents a compilation of challenges we encountered throughout the project. While they are not arranged in a specific order, they implicitly follow the sequence of milestones from one to four.

\begin{itemize}
\item Balancing the trade-off between the ease of prototyping in Python and its slower execution speed.
\item Working with Object-Oriented Programming (OOP) can be challenging, particularly when the software design lacks clarity.
\item The process of writing and debugging the chess engine was challenging. A more structured approach would have been preferable to trial and error.
\item Early unit testing proved to be less beneficial than anticipated.
\item The initial phase was the most difficult, as we were uncertain about where to begin.
\item The indiscriminate use of numpy does not automatically enhance code performance and can, in fact, slow it down.
\item Dealing with extensive functions with large logical statements.
\item Implementing advanced AI algorithms, such as Principal Variation Search (PVS) and Monte Carlo Tree Search (MCTS), in a way that effectively improved the AI's performance.
\item Dealing with the complexities of implementing a transposition table, including handling hash collisions.
\item Ensuring the accuracy and efficiency of the AI's evaluation process, particularly when introducing new features.
\item Managing the project's scope and complexity, especially when adding new features or making significant changes to the AI or backend.
\item Overcoming the limitations of the initial AI framework and restructuring it to accommodate new features.
\item Making multiple changes simultaneously, as opposed to implementing atomic changes.
\item Committing to tasks that ultimately did not yield significant benefits.
\item Coordinating team efforts, primarily due to the fact that the engine could effectively be worked on by only one person at a time.
\item Navigating and resolving GitHub merge conflicts among team members, which required a clear understanding of the code changes made by each individual and effective communication to ensure that important updates were not inadvertently discarded.
\end{itemize}

Despite these challenges, most provided valuable learning opportunities. The lessons we gleaned from this project will be discussed in the subsequent chapter.
